---
title: "Fitting_Manaus_wave2"
author: "Jennie Lavine"
date: "3/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(deSolve)
require(tidyverse)
require(bbmle)
```

Read in data on background death rates by age.

```{r}

#data from https://transparencia.registrocivil.org.br/dados-covid-download

obitos.2019<-read_csv('manaus_data/obitos-2019.csv')
obitos.2020<-read_csv('manaus_data/obitos-2020.csv')
obitos.2021<-read_csv('manaus_data/obitos-2021.csv')

obitos.2019$year<-2019
obitos.2020$year<-2020
obitos.2021$year<-2021

obitos<-rbind(obitos.2019,obitos.2020,obitos.2021)
obitos.am <-subset(obitos, uf=='AM')

res <- as.data.frame(
  cbind(
    year=rep(
      2019:2021,
      each=length(unique(obitos.am$faixa_etaria))),
    age_grp=rep(
      unique(obitos.am$faixa_etaria), 
      times=3), 
    deaths=0,
    props=0))

res$age_grp<-ordered(res$age_grp, levels = unique(res$age_grp)[1:12])

res$deaths<-as.numeric(res$deaths)
res$props<-as.numeric(res$props)

for(i in unique(res$year)){
  for(j in unique(res$age_grp)){
   row.ind<-which(res$year==i & res$age_grp==j)
   res[row.ind,'deaths'] <-
     sum(obitos.am$total[
       obitos.am$year==i & 
         obitos.am$faixa_etaria==j])
  }
  }

for(i in 2019:2021){
  res$props[res$year==i] <- 
    res$deaths[res$year==i] /  
    sum(res$deaths[res$year==i])
}

```

Read in demographic data on the population of Manaus (age distribution, birth rates, IFR, etc)

```{r}
##Manaus demographics and IFR from first wave

#https://www.citypopulation.de/en/brazil/amazonas/manaus/130260305__manaus/

age_dist <- as.data.frame(
  cbind(max_age= c(seq(10,70, by=10), 100), 
        pop.size.2010=c(
          328431, 356046, 365847, 306638, 
          207900,  129071, 63695, 44386)))

age_dist$prop <- age_dist$pop.size.2010/sum(age_dist$pop.size.2010)

#this is the age distribution just for the city of manaus
#pop size of manaus 2020 taken from wikipedia
pop.2020 <- 2219580
#https://en.wikipedia.org/wiki/Amazonas_(Brazilian_state)#Demographics
amazonas.pop.2019 <- 4144597

age_dist$pop2020<- age_dist$prop * pop.2020

deaths.am <- subset(res, year==2019 & age_grp!='N/I')$deaths
#assuming the death rates are the same in manaus in particular as amazonas as a whole...
deaths.man <- deaths.am*(pop.2020/amazonas.pop.2019)
age_dist$deaths <- c(deaths.man[1:7], sum(deaths.man[8:11]))
age_dist$death.rate <- age_dist$deaths / age_dist$pop2020

plot(age_dist$max_age, age_dist$death.rate, pch=1, lty=1, type='b')

#extracted from buss supplement thru june
#first four are a bit fudged because values are too low to accurately extract graphically
ifr.manaus.raw <- c(0.00001, 0.00001, 0.00002, 0.0005, 0.0016, 0.0043, 0.014, 0.03, 0.062)

#average last two IFRs to match number of age classes
ifr.manaus <- c(0.00001, 0.00001, 0.00002, 0.0005, 0.0016, 0.0043, 0.014, mean(c(0.03, 0.062)))

age_dist$ifr.manaus <- ifr.manaus


points(age_dist$max_age, age_dist$ifr.manaus, col=2, type='b', pch=2, lty=2)
legend('topleft', legend=c('background death rate','IFR'), pch=1:2, lty=1:2, col=1:2)
#compare to verity: 0.000016, 0.0000695, 0.000309, 0.000844, 0.00161, 0.00595, 0.0193, 0.0428, 0.078


#https://g1.globo.com/am/amazonas/noticia/2020/12/11/amazonas-lidera-ranking-de-registros-de-nascimento-tardios-no-pais-aponta-ibge.ghtml
#80997 births recorded in 2019.  8853 of them were from a previous year, so there is clearly error, but on the assumption that there are an equal number of births that occurred in 2019 that were NOT recorded, we use that number as it stands.
tot_pop <- sum(age_dist$pop2020)
birth.rate <- 80997/tot_pop
```

Read in data on excess deaths in Manaus in 2020 compared with 2019.

```{r}
#excess deaths data
#https://docs.google.com/spreadsheets/d/1UMhKX4mBSBM8YCZL338vB0UY8TYsgGTb/edit#gid=1271932269

burial.dat <- read_csv('manaus_data/excess_deaths_burials.csv')
colnames(burial.dat)[10] <- 'pub_priv_burials_2020_21'
colnames(burial.dat)[12] <- 'pub_priv_burials_cremats_2020_21'
burial.dat <- burial.dat[,1:27]
burial.dat$pub_priv_burials_cremats_2020_21<- 
  as.numeric(burial.dat$pub_priv_burials_cremats_2020_21)
burial.dat$Covid_dates_2020_2021 <- 
  as.Date(burial.dat$Covid_dates_2020_2021, '%d-%b-%y')

burial.dat$PreCovid_adj_deaths_15d_running_mean <-
  as.numeric(burial.dat$PreCovid_adj_deaths_15d_running_mean)

burial.dat<-subset(burial.dat, 
                   Covid_dates_2020_2021<=as.Date('2021-03-01') & 
                     Covid_dates_2020_2021>=as.Date('2020-04-01'))

diff.death <- 
  burial.dat$pub_priv_burials_cremats_2020_21- 
  burial.dat$PreCovid_adj_deaths_15d_running_mean

diff.death[is.na(diff.death)]<-0

plot(burial.dat$PreCovid_natual_deaths_ARPEN)
#points(table(am_dat_2019$date), type='p', col=2)
plot(burial.dat$pub_priv_burials_cremats_2020_21)

plot(burial.dat$Covid_dates_2020_2021, diff.death, 
     xaxt='n', xlab='Date', ylab='Excess deaths')
r <- range(burial.dat$Covid_dates_2020_2021)
axis.Date(1, at = seq(r[1], r[2], by = "month"), format = "%b '%y")


plot(burial.dat$Covid_dates_2020_2021,  cumsum(diff.death), 
     xaxt='n', xlab='Date', ylab='Cumulative excess deaths')
r <- range(burial.dat$Covid_dates_2020_2021)
axis.Date(1, at = seq(r[1], r[2], by = "month"), format = "%b '%y")


```

Set up model with Gamma-distributed durations of immunity.

```{r}
#Run model to show effects of various parameters on timing and size of second wave

covid.eq.ages.fun <- function (shape, nageclass){
  eqnames <- c('dS1', 'dI1', 'dS2', 'dI2', 'dS3', 'dI3', 
               paste('dRp',1:shape, sep=''), 
               paste('dRs',1:shape, sep=''))
  statenames <-c('S1', 'I1', 'S2', 'I2', 'S3', 'I3', 
               paste('Rp',1:shape, sep=''), 
               paste('Rs',1:shape, sep=''))
  eqs <- list(NULL)
  
  eqs[[1]]<- quote(dS1 <- mu * N + age.in * S1 - 
                     (lambda + delta + age.out + nu) * S1)
  
  eqs[[2]] <- quote(dI1 <- age.in * I1 + 
                      lambda *  S1 - 
                      (gamma + delta + age.out + alpha) * I1)
  
  
  eqs[[3]] <- parse(text = paste('dS2 <- age.in * S2 + omega * shape *' , 
                                 statenames[tail(grep('Rp', statenames),1)], 
                                 '-(lambda + delta + age.out + nu) * S2'))[[1]]
  
  eqs[[4]] <- quote(dI2 <- age.in * I2 + 
                      lambda * S2 -
                      (gamma + delta + age.out + alpha) * I2)
  
  eqs[[5]] <- parse(text = paste('dS3 <- age.in * S3 + omega * shape *' , 
                                 statenames[tail(grep('Rs', statenames),1)], 
                                 '-(lambda + delta + age.out + nu) * S3'))[[1]]
  
  eqs[[6]] <- quote(dI3 <- age.in * I3 + 
                      lambda * S3 -
                      (gamma + delta + age.out + alpha) * I3)
  
  eqs[[7]] <- quote(dRp1 <- age.in*Rp1 +  gamma * I1 - 
                      (omega * shape + delta + age.out) * Rp1)
  
    if(shape>1){
    for(i in 2:shape){
      eqs[[6+i]] <- parse(text = paste(eqnames[6+i],
                                       '<- age.in *', statenames[6+i],
                                       '+ omega * shape *', statenames[5+i], 
                                       ' - (delta + omega * shape + age.out) *', 
                                       statenames[6+i]) )[[1]]
    }
    } 
  eqs[[7+shape]] <- quote(dRs1 <- age.in*Rs1 +  gamma * (I2+I3) - 
                      (omega * shape + delta + age.out) * Rs1)
  
  if(shape>1){
    for(i in 2:shape){
      eqs[[6+shape+i]] <- parse(text = paste(eqnames[6+shape+i],
                                       '<- age.in *', statenames[6+shape+i],
                                       '+ omega * shape *', statenames[5+shape+i], 
                                       ' - (delta + omega * shape + age.out) *', 
                                       statenames[6+shape+i]) )[[1]]
    }
  
  }
  return(list(eqs=eqs, snames = statenames, enames = eqnames, shape=shape))
}

MODEL <- function(t, logstate, parameters, eqs) {
  state=exp(logstate)
  with(as.list(c(state, parameters)), {
    S1 = matrix(state[1:J], nrow = J, ncol=1)
    I1 = matrix(state[(J+1):(2*J)], nrow = J, ncol=1)
    S2 = matrix(state[(2*J+1):(3*J)], nrow = J, ncol=1)
    I2 = matrix(state[(3*J+1):(4*J)], nrow = J, ncol=1)
    S3 = matrix(state[(4*J+1):(5*J)], nrow = J, ncol=1)
    I3 = matrix(state[(5*J+1):(6*J)], nrow = J, ncol=1)
    Rp1 = matrix(state[(6*J+1):(7*J)], nrow = J, ncol=1)
    if(eqs$shape>1){
      for(i in 8:(6+eqs$shape)){
        eval(parse(
          text = paste(eqs$snames[i],
                       '<- matrix(state[(', i-1,
                       '*J+1):(', i, '*J)], nrow=J,  ncol=1)',
                       sep='')
        ) [[1]])
      }
    }
    Rs1 = matrix(state[((6+eqs$shape)*J+1):((7+eqs$shape)*J)], nrow = J, ncol=1)
    if(eqs$shape>1){
      for(i in (8+eqs$shape):(6+(2*eqs$shape))){
        eval(parse(
          text = paste(eqs$snames[i],
                       '<- matrix(state[(', i-1,
                       '*J+1):(', i, '*J)], nrow=J,  ncol=1)',
                       sep='')
        ) [[1]])
      }
    }
    
    N = sum(state)
    beta = beta0*(1+beta1*cos(2*pi*t+(pi/3)))
    age.in <- matrix(c(0, aging), nrow=J, ncol=1)
    age.out <- matrix(c(aging, 0), nrow=J, ncol=1)
    lambda = beta * sum(I1 + I2*rho + I3*rho)
    shape = eqs$shape
    
    res<-rep(0,length=length(eqs$enames)*J)
    names(res)<-names(state)
    for(i in 1:length(eqs$eqs)){
      res[(J*(i-1)+1):(J*i)]<- eval(eqs$eqs[[i]])
    }
    return(list(res/state))
  }) 
}
```

Set the Gamma shape parameter to 9 and the number of age classes to 8 (to match the data).
Write a function to transform the age-specific infection fatailty ratio.  Set initial conditions and fixed parameters.
```{r}
gshape=9
J=8

ifr.trans.fun <- function(x, psi){
  #if(is.na(psi)){return(x)} else
  #return(((1+(1/psi))*x)/(x+(1/psi)))
  return(x*psi)
}

test.eqs<-covid.eq.ages.fun(shape=gshape)
inits <- rep(0.0000000001,length(test.eqs$snames)*J)
state.names<-paste(rep(test.eqs$snames,each=J),
                   rep(letters[1:J]),sep='')
names(inits)<-state.names

inits[grep('S1',state.names)]<-age_dist$prop*0.3
inits[grep('Rp1',state.names)]<-age_dist$prop*0.7
inits[grep('I1', state.names)]<-1/sum(age_dist$pop.size)
inits<-log(inits/sum(inits))

params <- list(
  mu = matrix(c(birth.rate, rep(0, (J-1))), nrow=J, ncol=1),
  delta = matrix(age_dist$death.rate, nrow=J, ncol=1),
  alpha=0,
  beta0 = 250,
  beta1 = 0,
  aging = matrix(1/(diff(age_dist$max_age)),nrow=(J-1), ncol=1),
  gamma = 365/9,
  omega = 2,
  rho = 0.5,
  nu = 0,
  psi1=2,
  psi2=0
)

#deaths_res_tot <- matrix(NA, nrow=length(times)-1, ncol=nrow(param_grid_temp))

dates<-burial.dat$Covid_dates_2020_2021
intro.date<-as.Date('2020-11-06')
pulse.date<-as.Date('2020-05-01')
data <- diff.death[dates>intro.date]
max.date<-max(dates)
dates.sim <- dates[dates>pulse.date]
dates.fit <- seq(intro.date, max.date, by=1)

steps.yr = 365
maxt = 0.5
times1 = seq(0, by=1/steps.yr, length=intro.date-pulse.date)
times2 = seq(0, by=1/steps.yr, length=max.date-intro.date+1)
```

Write a function to estimate total deaths in the second wave using the following steps:
1. Simulate from the model with the initial conditions of 70\% immune on May 1, 2020, age distribution of immunity and susceptibility assumed to be proportional to the age distribution of the population.  Let this simulation run until the estimated introduction of the P.1 variant (Nov 6, 2021).
2. On Nov 6, set the initial conditions to the tail of the previous simulation, change the $R_0$ to consider the possibility of a higher $R_0$ for the P.1 variant, and simulate through the end of February.
3. Calculate the model-predicted deaths in the second wave (Nov 6, 2020 - Mar 1, 2021).


```{r}
sim.fun <- function(x=inits, parms=params,t1=times1, t2=times2, beta0=NA, omega=NA, psi1=NA, psi2=NA){
  if(!is.na(beta0)){params$beta0 <- beta0}
  if(!is.na(omega)){params$omega <- omega}
  if(!is.na(psi1)){params$psi1 <- psi1}
  if(!is.na(psi2)){params$psi2 <- psi2}
  params1<-params
  params1$beta0<-3*(params$gamma + params$mu)

  out1 <- ode(
    y = inits,
    t = t1,
    func = MODEL,
    parms = params1,
    eqs = test.eqs
  )
  
  inits2<-c(tail(out1[,-1], 1))
  names(inits2)<-names(inits)
  
  out2 <- ode(
    y = inits2,
    t = t2,
    func = MODEL,
    parms = params,
    eqs = test.eqs)
  
  temp <- exp(rbind(out1, out2))
  temp[,1] <- log(temp[,1])
  temp<-temp[dates.sim>=intro.date,]
  
  I1 <- temp[,grep('I1', colnames(temp))] 
  I2 <- temp[,grep('I2', colnames(temp))]
  
  probs.step <- 1-pexp(rate=as.numeric(params$gamma), 1/steps.yr)
  
  I1_t1 <- tail(I1, -1)
  I1_t0 <- head(I1, -1)
  X = I1_t1 - (I1_t0 * probs.step)  #calculate X, new infections at time t
  newI1_step<-X*tot_pop
  
  I2_t1 <- tail(I2, -1)
  I2_t0 <- head(I2, -1)
  X = I2_t1 - (I2_t0 * probs.step)  #calculate X, new infections at time t
  newI2_step<-X*tot_pop
  
  deaths_res_1 <- 
    apply(t(t(newI1_step) * 
              ifr.trans.fun(
                x=age_dist$ifr.manaus,
                psi=params$psi1)), 1, sum)
  
  
  deaths_res_2 <-
    apply(t(t(newI2_step) *
              ifr.trans.fun(x=age_dist$ifr.manaus,
                            psi=params$psi2)), 1, sum)
  
  deaths_res_tot <- deaths_res_1 + deaths_res_2
  return(deaths_res_tot)
}


```

Write a function (simLL) to calculate the likelihood of the data given a model with fixed parameters then find the maximum likelihood estimate for $\omega$ (waning of immunity), $\beta_0$ (transmissibility of new variant), and $\psi_1$, the fold-increase in the age-specific IFR for the new variant.  This model assumes that one infection provides complete protection against death upon a second infection (i.e., $psi_2$ is fixed at 0).

```{r,eval=F}
simLL <- function(psi1, omega, beta0) {
    x0 <- inits
    out <- sim.fun(psi1=psi1, omega=omega, beta0=beta0)
    SD <- sqrt(sum( (data-out)^2)/(length(data)))
    -sum(dnorm(data, mean=out, sd=SD, log=TRUE))
}
 
lower.pars<- c(psi1=1, omega=0, beta0=1.5*(365/9 + 1/52))  
fit <- mle2(simLL, 
            start=list(psi1=3, omega=3, beta0=5*(365/9 + 1/52)),  
            #method='L-BFGS-B',
            method='Nelder-Mead',
            #lower=lower.pars,
            control=list(maxit=1E5,trace=2),
            trace=T)
```

```{r, eval=F}
theta <- as.numeric(coef(fit))
best.sim<-sim.fun(psi1=theta[1], omega=theta[2], beta0=theta[3])
plot(dates.fit[-1], best.sim, type='l')
points(dates.fit[-1], data, col=2)
```


fit <- mle2(simLL, 
            start=list(psi1=4, omega=1.5, beta0=7*(365/9 + 1/52)),  
            method='L-BFGS-B',
            lower=lower.pars,
            control=list(maxit=1E5,trace=2),
            trace=T)
            
Coefficients:
      Estimate Std. Error z value Pr(z)
psi1    4.0745         NA      NA    NA
omega   0.0000         NA      NA    NA
beta0 311.0678         NA      NA    NA

-2 log L: 1048.963 







Call:
mle2(minuslogl = simLL, start = list(psi1 = 3, omega = 3, beta0 = 5 * 
    (365/9 + 1/52)), method = "L-BFGS-B", trace = T, lower = lower.pars, 
    control = list(maxit = 1e+05, trace = 2))

Coefficients:
        Estimate Std. Error z value     Pr(z)    
psi1    5.699281   0.122608  46.484 < 2.2e-16 ***
omega   2.346675   0.012219 192.058 < 2.2e-16 ***
beta0 123.562794   0.760960 162.377 < 2.2e-16 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

-2 log L: 955.065 


Same as function above, but this time we assume that immunity against transmission and pathology wane at the same time, so secondary infections are as deadly as primary infections ($psi_1 = psi_2$)

```{r, eval=F}
simLL.ifr2 <- function(ifr.mult, omega, beta0) {
    x0 <- inits
    out <- sim.fun(psi1=ifr.mult, psi2=ifr.mult, omega=omega, beta0=beta0)
    SD <- sqrt(sum( (data-out)^2)/(length(data)))
    -sum(dnorm(data, mean=out, sd=SD, log=TRUE))
}
 
lower.pars<- c(ifr.mult=0, omega=0, beta0=1.5*(365/9 + 1/52))  

fit.ifr2 <- mle2(simLL.ifr2, 
            start=list(ifr.mult=1.4, omega=1.9, beta0=3.95*(365/9 + 1/52)),  
            method='L-BFGS-B',
            lower=lower.pars,
            control=list(maxit=1E5,trace=2),
            trace=T)
```

```{r, eval=F}
theta <- as.numeric(coef(fit.ifr2))
best.sim<-sim.fun(psi1=theta[1], psi2=theta[1], omega=theta[2], beta0=theta[3])
plot(dates.fit[-1], best.sim)
points(dates.fit[-1], data, col=2)
```

    Call:
mle2(minuslogl = simLL, start = list(ifr.mult = 1.4, omega = 1.9, 
    beta0 = 3.95 * (365/9 + 1/52)), method = "L-BFGS-B", trace = T, 
    lower = lower.pars, control = list(maxit = 1e+05, trace = 2))

Coefficients:
           Estimate Std. Error z value     Pr(z)    
ifr.mult   1.726813   0.036297  47.575 < 2.2e-16 ***
omega      2.343298   0.011988 195.478 < 2.2e-16 ***
beta0    124.068841   0.757334 163.823 < 2.2e-16 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

-2 log L: 953.0351 

Plot results

```{r}
#Summary plot of 3 plausible scenarios
#1. 

sim <- function(inits, params, t1, t2){
  params1<-params
  params1$beta0<-3*(params$gamma + params$mu)
out1 <- ode(
    y = inits,
    t = t1,
    func = MODEL,
    parms = params1,
    eqs = test.eqs
  )

inits2<-c(tail(out1[,-1], 1))
names(inits2)<-names(inits)

out2 <- ode(
     y = inits2,
    t = t2,
    func = MODEL,
    parms = params,
    eqs = test.eqs)

  temp <- exp(rbind(out1, out2))
  temp[,1] <- log(temp[,1])

  I1 <- temp[,grep('I1', colnames(temp))] 
  I2 <- temp[,grep('I2', colnames(temp))]
  
  probs.step <- 1-pexp(rate=as.numeric(params$gamma), 1/steps.yr)
  
  I1_t1 <- tail(I1, -1)
  I1_t0 <- head(I1, -1)
  X = I1_t1 - (I1_t0 * probs.step)  #calculate X, new infections at time t
  newI1_step<-X*tot_pop
  
  I2_t1 <- tail(I2, -1)
  I2_t0 <- head(I2, -1)
  X = I2_t1 - (I2_t0 * probs.step)  #calculate X, new infections at time t
  newI2_step<-X*tot_pop
  
  deaths_res_1 <- 
     apply(t(t(newI1_step) * 
               ifr.trans.fun(
                 x=age_dist$ifr.manaus,
                 psi=params$psi1)), 1, sum)
  

   deaths_res_2 <-
   apply(t(t(newI2_step) *
                ifr.trans.fun(x=age_dist$ifr.manaus,
                              psi=params$psi2)), 1, sum)

  deaths_res_tot <- deaths_res_1 + deaths_res_2
  return(list(I1=apply(newI1_step, 1, sum), 
              I2=apply(newI2_step, 1, sum),
              D1 = deaths_res_1,
              D2 = deaths_res_2))
}


omega.vec<-c(2.34, 2.35)
psi1.vec<-c(1.7, 5.7)
psi2.vec<-c(1.7, 0)
beta0.vec<-c(124, 124)
#best.pars<-rbind(omega.vec, psi1.vec, psi2.vec, beta0.vec)
best.sims<-list(NULL)

params <- list(
  mu = matrix(c(birth.rate, rep(0, (J-1))), nrow=J, ncol=1),
  delta = matrix(age_dist$death.rate, nrow=J, ncol=1),
  alpha=0,
  beta0 = 250,
  beta1 = 0,
  aging = matrix(1/(diff(age_dist$max_age)),nrow=(J-1), ncol=1),
  gamma = 365/9,
  omega = 2,
  rho = 0.5,
  nu = 0,
  psi1=2,
  psi2=0
)



for(i in 1:2){
  params_temp<-params
  params_temp$omega<-omega.vec[i]
  params_temp$psi1 <- psi1.vec[i]
  params_temp$psi2 <- psi2.vec[i]
  params_temp$beta0<-beta0.vec[i]
  best.sims[[i]]<-sim(params=params_temp, inits, t1=times1, t2=times2)
}

prop.D1<-sum(best.sims[[1]]$D1)/sum(best.sims[[1]]$D1, best.sims[[1]]$D2)

ifr1.vec <- NA
ifr2.vec <- NA
  
plot(dates, diff.death, col='darkgray', ylab='Daily deaths', xlab='Months 2020-21')
for(i in 1:2){
  temp<-best.sims[[i]]$D1 + best.sims[[i]]$D2
  points(dates.fit, temp[dates.sim>=intro.date], type='l', col=i+1, lwd=1)
  ifr1.vec[i] <- sum(best.sims[[i]]$D1) / sum(best.sims[[i]]$I1)
  ifr2.vec[i] <- sum(best.sims[[i]]$D2) / sum(best.sims[[i]]$I2)
}

legend('top', pch=c(1,rep(NA, 2)), lty=c(0, 1, 1), col=c('darkgray',2, 3),
       legend=c('data', 
                paste('ifr2 = ifr1 = ', signif(ifr1.vec[1],1), sep=''),
                paste('ifr2 = 0; ifr1 = ',signif(ifr1.vec[2],1), sep='')))

```


```{r}
#plotting projections
end.date<-as.Date('2025-12-31')
dates.sim.long = seq(pulse.date, end.date, by=1)
times1.long = seq(0, by=1/steps.yr, length=intro.date-pulse.date)
times2.long = seq(0, by=1/steps.yr, length=end.date-intro.date+1)



sim.fun.long <- function(x=inits, parms=params,t1=times1.long, t2=times2.long, beta0=NA, omega=NA, psi1=NA, psi2=NA, psi3=NA){
  if(!is.na(beta0)){params$beta0 <- beta0}
  if(!is.na(omega)){params$omega <- omega}
  if(!is.na(psi1)){params$psi1 <- psi1}
  if(!is.na(psi2)){params$psi2 <- psi2}
  params1<-params
  params1$beta0<-3*(params$gamma + params$mu)

  out1 <- ode(
    y = inits,
    t = t1,
    func = MODEL,
    parms = params1,
    eqs = test.eqs
  )
  
  inits2<-c(tail(out1[,-1], 1))
  names(inits2)<-names(inits)
  
  out2 <- ode(
    y = inits2,
    t = t2,
    func = MODEL,
    parms = params,
    eqs = test.eqs)
  
  temp <- exp(rbind(out1, out2))
  temp[,1] <- log(temp[,1])
  temp<-temp[dates.sim.long>=intro.date,]
  
  I1 <- temp[,grep('I1', colnames(temp))] 
  I2 <- temp[,grep('I2', colnames(temp))]
  I3 <- temp[,grep('I3', colnames(temp))]

  probs.step <- 1-pexp(rate=as.numeric(params$gamma), 1/steps.yr)
  
  I1_t1 <- tail(I1, -1)
  I1_t0 <- head(I1, -1)
  X = I1_t1 - (I1_t0 * probs.step)  #calculate X, new infections at time t
  newI1_step<-X*tot_pop
  
  I2_t1 <- tail(I2, -1)
  I2_t0 <- head(I2, -1)
  X = I2_t1 - (I2_t0 * probs.step)  #calculate X, new infections at time t
  newI2_step<-X*tot_pop
  
  I3_t1 <- tail(I3, -1)
  I3_t0 <- head(I3, -1)
  X = I3_t1 - (I3_t0 * probs.step)  #calculate X, new infections at time t
  newI3_step<-X*tot_pop
  
  deaths_res_1 <- 
    apply(t(t(newI1_step) * 
              ifr.trans.fun(
                x=age_dist$ifr.manaus,
                psi=params$psi1)), 1, sum)
  
  
  deaths_res_2 <-
    apply(t(t(newI2_step) *
              ifr.trans.fun(x=age_dist$ifr.manaus,
                            psi=params$psi2)), 1, sum)
  
  deaths_res_3 <-
    apply(t(t(newI3_step) *
              ifr.trans.fun(x=age_dist$ifr.manaus,
                            psi=params$psi2)), 1, sum)
  
  deaths_res_tot <- deaths_res_1 + deaths_res_2 + deaths_res_3
  return(deaths_res_tot)
}

best.sims.long<-list(NULL)
for(i in 1:length(omega.vec)){
  best.sims.long[[i]]<-sim.fun.long(psi1=psi1.vec[i], psi2=psi2.vec[i], psi3=0, omega=omega.vec[i], beta0=beta0.vec[i], t1=times1.long, t2=times2.long)
}

plot(dates.sim.long, c(data, rep(NA, length(dates.sim.long)-length(data))), col='darkgray', ylab='Daily deaths', xlab='Year')

for(i in 1:length(omega.vec)){
  points(head(dates.sim.long,length(best.sims.long[[i]])), best.sims.long[[i]], type='l', col=i+1)
}
legend('topright', pch=c(1,rep(NA, 2)), lty=c(0, rep(1, 2)), col=c('darkgray', 2:3),
       legend=c('data',paste('ifr2 = ifr1 = ', signif(ifr1.vec[1],1), sep=''),
                paste('ifr2 = 0; ifr1 = ',signif(ifr1.vec[2],1), sep='')))

```


```{r}
#comparing model likelihoods
aic.fits <- c(aic.psi2.0 = 955.065 + 6, 
              aic.psi = 953.0351 + 6)
aic.min<-min(aic.fits)
b<-exp((aic.min-aic.fits)/2)
aic.weight <- b/sum(b)
aic.weight

```

